1.5: Data Wrangling in the tidyverse

### 2019-09-10 static/slides/05-slides.Rmd
Finalize 5 slides and 5 practice

---
title: "1.5: Data Wrangling in the tidyverse"
subtitle: "ECON 480 · Econometrics · Fall 2019"
author: 'Ryan Safner<br> Assistant Professor of Economics <br> <a href="mailto:safner@hood.edu"><i class="fa fa-paper-plane fa-fw"></i> safner@hood.edu</a> <br> <a href="https://github.com/ryansafner/metricsf19"><i class="fa fa-github fa-fw"></i> ryansafner/metricsf19</a><br> <a href="https://metricsF19.classes.ryansafner.com"> <i class="fa fa-globe fa-fw"></i> metricsF19.classes.ryansafner.com</a><br>'
#date:
output:
  xaringan::moon_reader:
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML" # rescales math with css changes https://github.com/yihui/xaringan/issues/143
    lib_dir: libs
    #seal: false
    css: [custom.css, custom-fonts.css, "hygge"] #, metropolis, metropolis-fonts
    nature:
      beforeInit: ["macros.js", "https://platform.twitter.com/widgets.js"] # first is for rescaling images , second is for embedding tweets, https://github.com/yihui/xaringan/issues/100
      highlightStyle: solarized-dark #ocean rainbow #monokai rainbow solarized-light #tomorrow-night -blue -eighties zenburn
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc(%current% / %total% * 100%);">
          </div>
        </div>
    includes:
      in_header: header.html # for font awesome, used in title  
---

@@

@@

- Most real world data analysis transforming chaos into order
    - Data strewn from multiple sources
    - Missing data
    - Data not in readable form

@@

# Workflow of a Data Scientist

.pull-left[
.center[.polaroid[![](https://www.dropbox.com/s/zke6m65wp3upouj/workflow.png?raw=1)]]
]

.pull-right[
- Consider Hadley Wickham's famous diagram from [R for Data Science](http://r4ds.had.co.nz)
1. .shoutcaps[Import] raw data from out there in the world
2. .shoutcaps[Tidy] it into a form that you can use to 
3. .shoutcaps[Explore] the data
4. .shoutcaps[Communicate] results to target audience
]

# Workflow of a Data Scientist

.pull-left[
.center[.polaroid[![](https://www.dropbox.com/s/zke6m65wp3upouj/workflow.png?raw=1)]]
]

@@

3. .shoutcaps[Explore] the data:
- .whisper[Transform] existing variables into more useful ones
- Make plots to .whisper[visualize] .onfire[relationships]
- Create .whisper[models] to analyze, predict, and interpret causal relationships

- 95% of the day-to-day work of data scientists, researchers, econometricians: ".hl[data wrangling]"
]

# Workflow of a Data Scientist

.pull-left[
.center[
![](https://www.dropbox.com/s/jd68ky17gb7pmu0/datawranglingnyt.png?raw=1)
]
]

.pull-right[
.center[
![](https://www.dropbox.com/s/8kzy823yubdv8gh/datawranglingnyt2.png?raw=1)
]
]

.center[
[New York Times](https://www.nytimes.com/2014/08/18/technology/for-big-data-scientists-hurdle-to-insights-is-janitor-work.html)
]

@@

# The tidyverse

@@

- Core packages include ones we've discussed before: `ggplot2`, `dplyr`, `magrittr` among several others (`tidy` `readr`, `purr`, `forcats`, `stringr`)
- Loading any `tidyverse` package loads `magrittr` (so you can use `%>%`)
- Learn more at [tidyverse.org](tidyverse.org)

![](https://www.dropbox.com/s/81ijxswbwqor0up/tidyverse.png?raw=1)

@@

- Easiest to just load the core tidyverse all at once
    - Note loading the tidyverse is "noisy", it will spew a lot of messages
    - Hide them with `suppressPackageStartupMessages()` and insert the `library()` command inside the `()`

@@

```{r, message=TRUE, warning=TRUE}
library("tidyverse")
```

@@

![](https://www.dropbox.com/s/9h1trto0iapnklx/tidyverse2.png?raw=1)

@@

@@

- We will make **extensive** use of:

1. `dplyr` for data wrangling
2. `tidyr` for tidying data
3. `ggplot2` for plotting data

- Loading `tidyverse` makes **automatically** uses

4. `tibble` for better data frames
5. `magrittr` for "pipeable" code

- We might later look at

6. `readr` for importing data
7. `forcats` for working with factors
8. `stringr` for working with strings

@@

# Tibble

@@

- `tibble` converts all `data.frames` into a .whisper[friendlier] version called `tibbles`

@@

# Tibble

@@

- Prints much nicer output in R
- Shows a bit of the `str`ucture of a dataframe automatically:
    - `nrow() x ncol()`
    - `<dbl>` is numeric ("double")
    - `<ord>` is an ordered factor
    - `<int>` is an integer
- Fundamental grammar of tidyverse: 
    1. start with a tibble 
    2. run a function on it
    3. output a new tibble
]

@@

# Magrittr I

@@

- The .hl[magrittr] package allows us to use the **"pipe" operator** (`%>%`)<sup>.red[1]</sup>
- `%>%` "pipes" the *output* of the *left* of the pipe *into* the *argument* of the *right* 
    - Running a function `f` on object `x` as `f(x)` becomes `x %>% f` in pipeable form
    - i.e. "take `x` and then run function `f` on it"
]

@@

.center[![](https://www.dropbox.com/s/74srdjrw13mgx4x/magrittr.png?raw=1)]
]

@@

# Magrittr II

@@

i.e. take `x` and perform function `f()` on `x` and then take that result and perform function `g()` on it
- With pipes, read operations from left $\rightarrow$ right:

@@

@@

.center[![](https://www.dropbox.com/s/74srdjrw13mgx4x/magrittr.png?raw=1)]
]

# Magrittr III

@@

@@

.center[![](https://www.dropbox.com/s/74srdjrw13mgx4x/magrittr.png?raw=1)]
]

@@

# Magrittr: Simple Example

@@

```{r}
# look at top 6 rows
head(gapminder) 
```
]

.pull-right[
```{r}
# use the pipe instead 
gapminder %>% head()
```
]

@@

# Magrittr: More Involved Example

- These two methods produce the same output (avg hwy mpg of Audi cars)

@@

# Readr

@@

- `Readr` lets you load common spreadsheet files (`.csv`, `.tsv`) with simple commands:
- `read_csv(path/to/my_data.csv)`<sup>.red[1]</sup>
    - `read_tsv(path/to/my_data.tsv)`
- Often this is enough, but there are many more customizations as necessary
- You can also export your data from R into a common spreadsheet file with:
- `write_csv(my_dataframe, path=path/to/file_name.csv)`<sup>.red[2]</sup> 
- Read more on the [tidyverse website](https://readr.tidyverse.org/) and the [Readr Cheatsheet](https://rawgit.com/rstudio/cheatsheets/master/data-import.pdf)
]

.footnote[<sup>.red[1]</sup> If you are using an R Project, this file path is in your main project folder, or any subfolders you decided to create, such as `data/my_data.csv`

<sup>.red[2]</sup> where `my_dataframe` is the name of the dataframe you want to export, and `file_name` is the name of the file you want to create.]

@@

.center[![](https://www.dropbox.com/s/rl1sj67y2m8fqy5/readxl.png?raw=1)]
.center[![](https://www.dropbox.com/s/ua6t5wz1se34vmo/haven.png?raw=1)]
]

@@

- Some data types, particularly for other software programs like Excel, STATA, SAS, and SPSS are not `.csv` or `.txt` spreadsheets
- `readxl` has commands equivalent to `readr`'s for Excel data types:
    - `read_xls("path/to/my/data.xls")` for older Excel `.xls` files
    - `read_xlsx("path/to/my/data.xlsx")` for modern Excel `.xls` files
    - `write_xls(my_dataframe, path=path/to/file_name.xls)` or ``write_xlsx(my_dataframe, path=path/to/file_name.xlsx)` to export as `xls` or `xlsx` files
- `haven` has commands equivalent to `readr`'s for these data types:
    - `read_dta("path/to/my_data.dta")` for STATA `.dta` files
    - `read_sav("path/to/my_data.sav")` for SPSS `.sav` files
    - `read_sas("path/to/my_data.sas7bdat")` for SAS files 
]

@@

- `dplyr` uses much more efficient and intuitive commands<sup>.magenta[1]</sup> to manipulate dataframes
- Grammatically:
    - `Base R` is all about passively running functions on nouns: `function(object)`
    - `dplyr` is all about actively using verbs: `dplyr_verb(df, conditions)`<sup>.red[2]</sup>
- Three great features:

1. Allows use of `%>%` pipe operator
2. Input and output is always a dataframe<sup>.magenta[3]</sup>
3. Shows the output from a manipulation, but does not save/overwrite as an object unless explicitly assigned to an object

@@

.footnote[<sup>.magenta[1]</sup> Particularly when combined with `magrittr`'s pipe (`%>%`) command!

<sup>.red[2]</sup> With the pipe, even simpler: `df %>% dplyr_verb(conditions)`

<sup>.red[3]</sup> Technically a `tibble` with `tidyverse`!]

@@

| Verb | Does | 
|----------|------|
| `filter` | Keep only selected *observations* | 
| `select` | Keep only selected *variables* |
| `arrange` | Reorder rows (e.g. in numerical order) |
| `mutate` | Create new variables | 
| `recode` | Change a variable's values or categories/`factor` levels |
| `summarize` | Collapse data into summary statistics| 
| `group_by` | Perform any of the above functions by groups/categories |

@@

# dplyr::filter

@@

```{r, eval=F}
# look only at African observations
# syntax without the pipe
filter(gapminder, continent=="Africa")

# using the pipe

@@

```{r, echo=F}
# syntax with the pipe 
gapminder %>%
  filter(continent=="Africa")
```

@@

```{r, eval = F}
# base syntax
africa <- filter(gapminder, continent=="Africa")

# using the pipe
africa <- gapminder %>%
  filter(continent == "Africa")

@@

```{r, eval = F}
# base syntax
gapminder <- filter(gapminder, continent=="Africa")

# using the pipe
gapminder <- gapminder %>%
  filter(continent == "Africa")

@@

# dplyr conditionals

- In many data wrangling contexts, you will want to select data .whisper[conditionally]
    - In computer code, you are selecting only observations for which a set of logical conditions are `TRUE`
- For example:
    - `>`, `<`: greater than, less than
    - `>=`, `<=`: greater than or equal to, less than or equal to
    - `==`, `!=`: is equal to, is not equal to
    - `%in%`: is a member of some defined set $(\in)$
    - `&`: AND<sup>1</sup>
    - `|`: OR

.footnote[<sup>1</sup> Commas also work in `dplyr`.]

@@

# dplyr::filter with conditionals

@@

```{r, eval=F}
# look only at African observations in 1997
gapminder %>%
  filter(continent == "Africa", year == 1997)
```
]

@@

# dplyr::filter with conditionals II

@@

```{r, eval=F}
# look only at African observations or observations in 1997
gapminder %>%
  filter(continent == "Africa" | year == 1997)
```
]

@@

# dplyr::filter with conditionals III

@@

```{r, eval=F}
# look only at U.S. and U.K. observations in 2002
gapminder %>%
  filter(country %in% c("United States", "United Kingdom"), year == 2002)
```
]

@@

# dplyr::arrange I

@@

```{r, eval=F}
# order by smallest to largest population
# syntax without the pipe
arrange(gapminder, pop)

# using the pipe

@@

# dplyr::arrange II

@@

```{r, eval=F}
# order by year, with the smallest to largest population in each year
# syntax without the pipe
arrange(gapminder, year, pop)

# using the pipe

@@

# dplyr::arrange III

@@

```{r, eval=F}
# order by largest to smallest population
# syntax without the pipe
arrange(gapminder, desc(pop))

# using the pipe

@@

# dplyr::select I

@@

```{r, eval=F}
# keep only country, year, and population variables
# syntax without the pipe
select(gapminder, country, year, pop)

# using the pipe

@@

# dplyr::select II

@@

```{r, eval=F}
# keep all except gdpPercap
# syntax without the pipe
select(gapminder, -gdpPercap)

# using the pipe

@@

# dplyr::select III

@@

```{r, eval=F}
# keep all and move pop first
# syntax without the pipe
select(gapminder, pop, everything())

# using the pipe

@@

# dplyr::select IV

@@

```{r}
# keep country and all variables containing "per"

@@

# dplyr::rename

- `rename` changes the name of a variable (column)
    - Format: `new_name=old_name`
--

@@

```{r, eval=F}
# rename gdpPercap to GDP
# syntax without the pipe
rename(gapminder, GDP = gdpPercap)

# using the pipe

@@

# dplyr::mutate

@@

# dplyr::mutate II

@@

```{r, eval=F}
# create variable "europe" if country is in Europe
# syntax without the pipe
mutate(gapminder, europe = ifelse(continent == "Europe", 
                                  yes = "In Europe", 
                                  no = "Not in Europe"))

# using the pipe

@@

```{r, echo=F}
# syntax with the pipe 
gapminder %>%
  mutate(europe = ifelse(continent == "Europe", 
                                  yes = "In Europe", 
                                  no = "Not in Europe"))
```
]

@@

# dplyr::mutate III

@@

```{r, eval=F}
# create population in millions 
# syntax without the pipe
mutate(gapminder, pop_millions = pop / 1000000)

# using the pipe

gapminder %>%
  rename(pop_millions = pop / 1000000)
```

@@

```{r, echo=F}
# syntax with the pipe 
gapminder %>%
  mutate(pop_millions = pop / 1000000)
```
]

@@

# dplyr::mutate IV

@@

```{r, eval=F}
# create GDP variable from gdpPercap and pop
# syntax without the pipe
mutate(gapminder, GDP = gdpPercap * pop)

# using the pipe

gapminder %>%
  rename(GDP = gdpPercap * pop)
```

@@

```{r, echo=F}
# syntax with the pipe 
gapminder %>%
  mutate(GDP = gdpPercap * pop)
```
]

@@

# dplyr::mutate V

@@

@@

@@

- Don't forget to assign the output to  new `tibble` if you want to "save" the new variables!

@@

# Exporting Tables with knitr::kable

>>>>

---
title: "1.5: Data Wrangling in the tidyverse"
subtitle: "ECON 480 · Econometrics · Fall 2019"
author: 'Ryan Safner<br> Assistant Professor of Economics <br> <a href="mailto:safner@hood.edu"><i class="fa fa-paper-plane fa-fw"></i> safner@hood.edu</a> <br> <a href="https://github.com/ryansafner/metricsf19"><i class="fa fa-github fa-fw"></i> ryansafner/metricsf19</a><br> <a href="https://metricsF19.classes.ryansafner.com"> <i class="fa fa-globe fa-fw"></i> metricsF19.classes.ryansafner.com</a><br>'
#date:
output:
  xaringan::moon_reader:
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML" # rescales math with css changes https://github.com/yihui/xaringan/issues/143
    lib_dir: libs
    #seal: false
    css: [custom.css, custom-fonts.css, "hygge"] #, metropolis, metropolis-fonts
    nature:
      beforeInit: ["macros.js", "https://platform.twitter.com/widgets.js"] # first is for rescaling images , second is for embedding tweets, https://github.com/yihui/xaringan/issues/100
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc(%current% / %total% * 100%);">
          </div>
        </div>
    includes:
      in_header: header.html # for font awesome, used in title  
---

@@

class: inverse, middle, center

### [tibble: friendlier dataframes](#12)

### [magrittr: piping code](#17)

### [readr: importing data](#26)

### [dplyr: wrangling data](#32)

### [dplyr::filter(): select observations](#35)

### [dplyr::arrange(): reorder observations](#50)
]

### [dplyr::select(): select variables](#60)

### [dplyr::rename(): rename variables](#73)

### [dplyr::mutate(): create new variables](#77)

### [dplyr::summarize(): create statistics](#98)

### [tidyr: reshaping data](#133)

### [dplyr: combining datasets](#134)

@@

- Most data analysis is taming chaos into order
    - Data strewn from multiple sources
    - Missing data ("`NA`")
    - Data not in a readable form

.center[
![:scale 90%](https://www.dropbox.com/s/n9fm2lkii5renzf/messyexcel.png?raw=1)
]
]

.pull-right[
.center[
![](https://www.dropbox.com/s/gp1lxciqaxq141q/datawrangling0.jpg?raw=1)
]
]
---

# Workflow of a Data Scientist I

.pull-left[
.center[
.polaroid[![](https://www.dropbox.com/s/zke6m65wp3upouj/workflow.png?raw=1)]

[R for Data Science](http://r4ds.had.co.nz)
]
]

.pull-right[
- Consider Hadley Wickham's famous diagram:
1. .whisper[Import] raw data from out there in the world
2. .whisper[Tidy] it into a form that you can use 
3. .whisper[Explore] the data (do these 3 repetitively!)
    - **Transform**
    - **Visualize**
    - **Model**
4. .whisper[Communicate] results to target audience

Ideally, you'd want to be able to do all of this in one program
]

@@

# Workflow of a Data Scientist II

.pull-left[
.center[
![](https://www.dropbox.com/s/jd68ky17gb7pmu0/datawranglingnyt.png?raw=1)
[New York Times](https://www.nytimes.com/2014/08/18/technology/for-big-data-scientists-hurdle-to-insights-is-janitor-work.html)
]

@@

> "Yet far too much handcrafted work - what data scientists call "**data wrangling**," "**data munging**," and "**data janitor work**" - is still required. Data scientists, according to interviews and expert estimates, spend from **50 to 80 percent of their time** mired in this more mundane labor of collecting and preparing unruly digital data, before it can be explored for useful nuggets."

@@

# The tidyverse I

@@

- Allows you to do all of those things with one (set of) package(s)!
- Learn more at [tidyverse.org](tidyverse.org)

.center[
![](https://www.dropbox.com/s/81ijxswbwqor0up/tidyverse.png?raw=1)
]

@@

- Easiest to just load the core tidyverse all at once
    - First install may take a few minutes - installs a lot of packages!
    - Note loading the tidyverse is "noisy", it will spew a lot of messages
    - Hide them with `suppressPackageStartupMessages()` and insert `library()` command inside

@@

```{r, message=TRUE, warning=TRUE}
# install for first time
# install.packages("tidyverse") # this takes a few minutes and may give several prompts

# load tidyverse 
suppressPackageStartupMessages(library("tidyverse"))
```

@@

- `tidyverse` contains a lot of packages, not all are loaded automatically

```{r}
tidyverse_packages()
```

@@

# Your Workflow in the tidyverse:

.center[
![:scale 80%](https://www.dropbox.com/s/9h1trto0iapnklx/tidyverse2.png?raw=1)
]
---

@@

- We will make **extensive** use of (and talk today about):

1. `tibble` for friendlier dataframes
2. `magrittr` for "pipeable" code
3. `readr` for importing data
4. `dplyr` for data wrangling
5. `tidyr` for tidying data
6. `ggplot2` for plotting data (we've already covered)

- We will (or might) later look at:
7. `broom` for tidy regression (not part of core tidyverse)
8. `forcats` for working with factors
9. `stringr` for working with strings
10. `lubridate` for working with dates and times
11. `purrr` for iteration

@@

class: inverse, center, middle
# tibble: friendlier dataframes

# tibble I

@@

- `tibble` converts all `data.frames` into a *friendlier* version called `tibbles` (or `tbl_df`)

@@

# tibble II

@@

- Prints much nicer output

- Shows a bit of the `str`ucture:
    - `nrow() x ncol()`
    - `<dbl>` is numeric ("double")
    - `<ord>` is an ordered factor
    - `<int>` is an integer

- Fundamental grammar of tidyverse: 
    1. start with a tibble 
    2. run a function on it
    3. output a new tibble
]

@@

# tibble III

@@

- Create a `tibble` from a `data.frame` with `as_tibble()`

```{r, eval=F}
as_tibble(mpg) # take built-in dataframe mpg
```

- Create a `tibble` from scratch with `tibble()`, works like `data.frame()`

```{r}
example<-tibble(x = seq(2,6,2), # sequence from 2 to 6 by 2's
                y = rnorm(3,0,1), # 3 random draws with mean 0, sd 1
                colors = c("orange", "green", "blue"))

example
```
]

# tibble IV

@@

- Create a `tibble` row-by-row with `tribble()`

```{r}
example_2<-tribble(
  ~x, ~y, ~color, # each variable name starts with ~
  2, 1.5, "orange",
  4, 0.2, "green",
  6, 0.8, "blue") # last element has no comma

example_2
```
]

class: inverse, center, middle
# magrittr: piping code

# magrittr I

.center[
![:scale 100%](https://www.dropbox.com/s/74srdjrw13mgx4x/magrittr.png?raw=1)
]
]

- The `magrittr` package allows us to use the **"pipe" operator** (`%>%`)<sup>.red[1]</sup>

- `%>%` "pipes" the *output* of the *left* of the pipe *into* the *(1<sup>st</sup>) argument* of the *right*

- Running a function `f` on object `x` as `f(x)` becomes `x %>% f` in pipeable form
    - i.e. "take `x` and then run function `f` on it"
]

@@

# magrittr II

.center[
![:scale 100%](https://www.dropbox.com/s/74srdjrw13mgx4x/magrittr.png?raw=1)
]
]

@@

i.e. take `x` and perform function `f()` on `x` and then take that result and perform function `g()` on it

- With pipes, read operations from left $\rightarrow$ right:

@@

# magrittr III

@@

.center[
![:scale 100%](https://www.dropbox.com/s/74srdjrw13mgx4x/magrittr.png?raw=1)
]
]

@@

# magrittr IV

@@

@@

# magrittr: Simple Example

@@

.pull-left[
```{r, eval=F}
# look at top 6 rows
head(gapminder)

# use pipe instead
gapminder %>% head()
```
]

```{r, echo=F}
gapminder %>% head()
```
]

@@

# magrittr: More Involved Example

- These two methods produce the same output (average hightway mpg of Audi cars)

@@

class: inverse, center, middle
# readr: importing data

# readr

@@

- `readr` helps load common spreadsheet files (`.csv`, `.tsv`) with simple commands:
- `read_*(path/to/my_data.*)`
    - where `*` can be `.csv` or `.tsv`

- Often this is enough, but many more customizations possible

- You can also *export* your data from R into a common spreadsheet file with:
- `write_*(my_df, path = path/to/file_name.*)`
    - where `my_df` is the name of your `tibble`, and `file_name` is the name of the file you want to save as

- Read more on the [tidyverse website](https://readr.tidyverse.org/) and the [Readr Cheatsheet](https://rawgit.com/rstudio/cheatsheets/master/data-import.pdf)
]

@@

.center[
![:scale 60%](https://www.dropbox.com/s/rl1sj67y2m8fqy5/readxl.png?raw=1)

![:scale 60%](https://www.dropbox.com/s/ua6t5wz1se34vmo/haven.png?raw=1)]
]

@@

- For other data types from software programs like Excel, STATA, SAS, and SPSS:
- `readxl` has equivalent commands for Excel data types:
    - `read_*("path/to/my/data.*")`
    - `write_*(my_dataframe, path=path/to/file_name.*)`
    - where `*` can be `.xls` or `.xlsx`

- `haven` has equivalent commands for other data types:
    - `read_*("path/to/my_data.dta")` for STATA `.dta` files
    - `write_*(my_dataframe, path=path/to/file_name.*)`
    - where `*` can be `.dta` (STATA), `.sav` (SPSS), `.sas7bdat` (SAS)
]

# Common Import Issues I

- Most common: *"where the hell is my data file"??*

- Recall `R` looks for files to `read_*()` in the default working directory (check what it is with `getwd()`, change it with `setwd()`)

- You can tell `R` where this data is by making the `path` a part of the file's name when importing
    - Use `..` to "move up one folder"
    - Use `/` to "enter a folder"

- Either use an **absolute path** on your computer:
```{r, eval=F}
# Example

df <- read_csv("C:/Documents and Settings/Ryan Safner/Downloads/my_data.csv")
```

- Or use a **relative path** *from* R's working directory
```{r, eval=F}
# Example
# If working directory is Documents, but data is in Downloads, like so: 
# 
# Ryan Safner
# |
# |- Documents/
# |- Downloads/
# |- Photos/
# |- Videos/
df <- read_csv("../Downloads/my_data.csv")
```

# Common Import Issues II

- Most common: *"where the hell is my data file"??*

- Recall `R` looks for files to `read_*()` in the default working directory (check what it is with `getwd()`, change it with `setwd()`)

- You can tell `R` where this data is by making the `path` a part of the file's name when importing
    - Use `..` to "move up one folder"
    - Use `/` to "enter a folder"

- Or use a **relative path** *from* R's working directory
```{r, eval=F}
# Example
# If working directory is Documents, but data is in Downloads, like so: 
# 
# Ryan Safner
# |
# |- Documents/
# |- Downloads/
# |- Photos/
# |- Videos/
df <- read_csv("../Downloads/my_data.csv")
```

# Common Import Issues III

- **Suggestion** to make your data import easier: *Download and move files to R's working directory*

- Your computer and working directory are different from mine (and others)

- This is *not* a reproducible workflow!

- We'll finally fix this next class with `R Projects`
    - The working directory is set to the Project Folder by default
    - Same for everyone on any computer!

class: inverse, center, middle
# dplyr: wrangling data

@@

- `dplyr` uses more efficient & intuitive commands to manipulate tibbles

- `Base R` grammar: passively running functions on nouns: `function(object)`
- `dplyr` grammar: actively using verbs: `dplyr_verb(df, conditions)`<sup>.red[1]</sup>

- Three great features:

1. Allows use of `%>%` pipe operator
2. Input and output is always a `tibble`
3. Shows the output from a manipulation, but does not save/overwrite as an object unless explicitly assigned to an object

@@

.footnote[<sup>.red[1]</sup> With the pipe, even simpler: `df %>% dplyr_verb(conditions)`]

@@

| Verb | Does | 
|----------|------|
| `filter()` | Keep only selected *observations* | 
| `select()` | Keep only selected *variables* |
| `arrange()` | Reorder rows (e.g. in numerical order) |
| `mutate()` | Create new variables | 
| `summarize()` | Collapse data into summary statistics| 
| `group_by()` | Perform any of the above functions by groups/categories |

@@

class: inverse, center, middle
# dplyr::filter(): select observations

.center[
![](https://www.dropbox.com/s/n9o9x29mlvyrr8q/filter.png?raw=1)
]

# dplyr::filter()

@@

```{r, eval=F}
# look only at African observations
# syntax without the pipe
filter(gapminder, continent=="Africa")
```

```{r, eval=F}
# using the pipe

@@

```{r, echo=F}
gapminder %>%
  filter(continent == "Africa")
```

@@

```{r, eval = F}
# base syntax
africa <- filter(gapminder,
                 continent=="Africa")
```

```{r, eval=F}
# using the pipe
africa <- gapminder %>%
  filter(continent == "Africa")
```

```{r, eval=F}
# look at new tibble
africa
```
]

@@

```{r, eval = F}
# base syntax
gapminder <- filter(gapminder,
                    continent=="Africa")
```

```{r, eval=F}
# using the pipe
gapminder <- gapminder %>%
  filter(continent == "Africa")

@@

# dplyr Conditionals

- In many data wrangling contexts, you will want to select data .whisper[conditionally]

- In computer code, you are selecting only observations for which a set of logical conditions are `TRUE`<sup>.red[1]</sup>

- For example:
    - `>`, `<`: greater than, less than
    - `>=`, `<=`: greater than or equal to, less than or equal to
    - `==`, `!=`: is equal to, is not equal to
    - `%in%`: is a member of some defined set $(\in)$
    - `&`: AND (commas also work instead)
    - `|`: OR
    - `!`: not

.footnote[<sup>.red[1]</sup> See `?Comparison` and `?Base::Logic`.]

@@

# dplyr::filter() with Conditionals

@@

```{r, eval=F}
# look only at African observations
# in 1997
gapminder %>%
  filter(continent == "Africa",
         year == 1997)
```
]

@@

# dplyr::filter() with Conditionals II

@@

```{r, eval=F}
# look only at African observations 
# or observations in 1997
gapminder %>%
  filter(continent == "Africa" |
           year == 1997)
```
]

@@

# dplyr::filter() with Conditionals III

@@

```{r, eval=F}
# look only at U.S. and U.K. 
# observations in 2002
gapminder %>%
  filter(country %in% 
           c("United States",
             "United Kingdom"),
         year == 2002)
```
]

@@

class: inverse, center, middle
# dplyr::arrange(): reorder observations

# dplyr::arrange() I

@@

```{r, eval=F}
# order by smallest to largest pop
# syntax without the pipe
arrange(gapminder, pop)
```

```{r, eval=F}
# using the pipe

@@

# dplyr::arrange() II

@@

```{r, eval=F}
# order by year, with the smallest 
# to largest pop in each year
# syntax without the pipe
arrange(gapminder, year, pop)
```

```{r, eval=F}
# using the pipe

@@

# dplyr::arrange() III

@@

```{r, eval=F}
# order by largest to smallest pop
# syntax without the pipe
arrange(gapminder, desc(pop))
```

```{r, eval=F}
# using the pipe

@@

class: inverse, center, middle
# dplyr::select(): select variables

.center[
![](https://www.dropbox.com/s/meq9tyvn4u5wcfb/select.png?raw=1)
]

# dplyr::select() I

@@

```{r, eval=F}
# keep only country, year, 
# and population variables
# syntax without the pipe
select(gapminder, country, year, pop)
```

```{r, eval=F}
# using the pipe

@@

# dplyr::select() II

@@

```{r, eval=F}
# keep all *except* gdpPercap
# syntax without the pipe
select(gapminder, -gdpPercap)
```

```{r, eval=F}
# using the pipe

@@

# dplyr::select() III

@@

```{r, eval=F}
# keep all and move pop first
# syntax without the pipe
select(gapminder, pop, everything())
```

```{r, eval=F}
# using the pipe

@@

# dplyr::select() IV

@@

```{r}
# keep country and all variables 
# containing "per"

@@

class: inverse, center, middle
# dplyr::rename(): rename variables

# dplyr::rename()

- `rename` changes the name of a variable (column)
    - Format: `new_name = old_name`
--

@@

```{r, eval=F}
# rename gdpPercap to GDP
# syntax without the pipe
rename(gapminder, GDP = gdpPercap)
```

```{r, eval=F}
# using the pipe

@@

class: inverse, center, middle
# dplyr::mutate(): create new variables

.center[
![](https://www.dropbox.com/s/1ktrsvya8pq3hh3/mutate.png?raw=1)
]

# dplyr::mutate()

@@

# dplyr::mutate() II

@@

```{r, eval=F}
# create variable "europe" if country 
# is in Europe
# syntax without the pipe
mutate(gapminder, 
       europe = ifelse(continent == "Europe",
                       yes = "In Europe",
                       no = "Not in Europe"))
```

```{r, eval=F}
# using the pipe

@@

```{r, echo=F}
# syntax with the pipe 
gapminder %>%
  select(1:3) %>% # hiding vars just for space constraints
  mutate(europe = ifelse(continent == "Europe",
                         yes = "In Europe",
                         no = "Not in Europe"))
```
]

@@

# dplyr::mutate() III

@@

```{r, eval=F}
# create population in millions 
# syntax without the pipe
mutate(gapminder,
       pop_mil = pop / 1000000)
```

```{r, eval=F}
# using the pipe

gapminder %>%
  rename(pop_mil = pop / 1000000)
```

@@

```{r, echo=F}
# syntax with the pipe 
gapminder %>%
  select(1:5) %>% # hiding vars just for space constraints
  mutate(pop_mil = pop / 1000000)
```
]

@@

# dplyr::mutate() IV

@@

```{r, eval=F}
# create GDP variable from gdpPercap 
# and pop, in billions
# syntax without the pipe
mutate(gapminder,
       GDP = ((gdpPercap * pop)/1000000000))
```

```{r, eval=F}
# using the pipe

gapminder %>%
  mutate(GDP = ((gdpPercap * pop)/1000000000))
```

@@

```{r, echo=F}
# syntax with the pipe 
gapminder %>%
  select(c(country, continent, year, pop, gdpPercap)) %>% # hiding vars just for space constraints
  mutate(GDP = ((gdpPercap * pop)/1000000000))
```
]

@@

# dplyr::mutate() V

- Change `class` of a variable inside `mutate()` with `as.*()`

```{r}
gapminder %>% head(., 2)

# change year from an integer to a factor
gapminder %>%
  mutate(year = as.factor(year))
```

# dplyr::mutate(): Multiple Variables

@@

# dplyr::transmute()

@@

# dplyr::mutate(): Conditionals

- Boolean, logical, and conditionals all work well in `mutate()`:

```{r}
gapminder %>%
  select(country, year, lifeExp) %>%
  mutate(long_1 = lifeExp > 70,
         long_2 = ifelse(lifeExp > 70, "Long", "Short"))
```

# dplyr::mutate(): order Aware

- `mutate()` is order-aware, so you can chain multiple mutates that depend on previous mutates

```{r}
gapminder %>%
  select(country, year, lifeExp) %>%
  mutate(dog_years = lifeExp * 7,
         comment = paste("Life expectancy in", country, "is", dog_years, "in dog years.", sep = " "))
```

# dplyr::mutate(): case_when()

- `case_when` creates a new variable with values that are conditional on values of other variables (e.g., "if/else")
    - Last argument: `TRUE`: when 
```{r}
gapminder %>%
  mutate(European = case_when(
    continent == "Europe" ~ "Aye",
    TRUE ~ "Nay"
  ))
```

# dplyr::mutate(): scoped I

- "Scoped" variants of `mutate` that work on a subset of variables:
    - `mutate_all()` affects every variable
    - `mutate_at()` affects named or selected variables
    - `mutate_if()` affects variables that meet a criteria

```{r}
# round all observations of numeric
# variables to 2 digits
gapminder %>%
  mutate_if(is.numeric, round, digits = 2)
```

# dplyr::mutate(): scoped II

- "Scoped" variants of `mutate` that work on a subset of variables:
    - `mutate_all()` affects every variable
    - `mutate_at()` affects named or selected variables
    - `mutate_if()` affects variables that meet a criteria

```{r}
# make all factor variables uppercase
gapminder %>%
  mutate_if(is.factor, toupper) 
```

# dplyr::mutate()

- Don't forget to assign the output to a new `tibble` (or overwrite original) if you want to "save" the new variables!

class: inverse, center, middle
# dplyr::summarize(): create statistics

.center[
![](https://www.dropbox.com/s/d7hiv0fst5nhhzu/summarize.png?raw=1)
]

# dplyr::summarize() I

- `summarize`<sup>.red[1]</sup> outputs a tibble of desired summary statistics
    - can name the statistic variable as if you were `mutate`-ing a new variable
--

```{r, eval=F}
# get average life expectancy
# call it avg_LE
summarize(gapminder,
          avg_LE = mean(lifeExp))
```

```{r, eval=F}
# using the pipe

gapminder %>%
  summarize(avg_LE = mean(lifeExp))
```

@@

```{r, echo=F}
gapminder %>%
  summarize(avg_LE = mean(lifeExp))
```

.footnote[<sup>.red[1]</sup> Also the more civilised non-U.S. English spelling `summarise` also works. `dplyr` was written by a Kiwi after all!
]
---

# dplyr::summarize() II

- Useful `summarize()` commands:

| Command | Does |
|---------|------|
| `n()`<sup>.red[*]</sup>   | Number of observations |
| `n_distinct()`<sup>.red[*]</sup> | Number of unique observations |
| `sum()` | Sum all observations of a variable |
| `mean()` | Average of all observations of a variable |
| `median()` | 50<sup>th</sup> percentile of all observations of a variable |
| `sd()`  | Standard deviation of all observations of a variable |

.footnote[<sup>.red[*]</sup> Most commands require you to put a variable name inside the command's argument parentheses. These commands require nothing to be in parentheses!]

@@

# dplyr::summarize() II

- Useful `summarize()` commands (continued):
